---
title: "Introduction to Reading Tabular Data in R"
output: html_document
---

```{r include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  eval = FALSE
  )

# echo = FALSE it will not share the code that you wrote to get to an output
# message = FALSE, does not print out messages associated with running code
# eval = FALSE, does not automatically run all of the code, if this is set = TRUE, you will see the result
```

# Reading Tabular Data in R

## Introduction

Data sets are stored in tabular format and there are many possible ways to organize tabular data. Some organizational schemes are designed to be easily read on the page (or screen), while others are designed to be easily used in analysis. In this tutorial, we focus on how a dataset stored in a tabular format can be read in to R. 

## Common file extensions

Here is a small list of common file extensions you may come across. 

| Format | Typical Extension | Ramler Recommended Package |
| ----------- | ------ | ---------------------------|
| Comma-separated data | .csv | `readr` (read_csv)|
| Tab-separated data | .tsv | `readr`  |
| Semi-colon separated data | .csv | `readr` |
| Pipe ("\|")  separated data | .psv | `readr` |
| Excel | .xls or .xlsx| `readxl` ...or just resave it as a csv |
| R syntax | .R | base R |
| Saved R objects | .RData, .rda | base R  |
| [Google Sheets](https://www.google.com/sheets/about/) | _URL_| `googlesheets4` or resave it as a comma-separated data  |
| JSON (a possible topic of the course)| .json | `jsonlite` |
| SQL Database  (a possible topic of the course) | _varies_ |`dbplyr` or one that is specific to the variety of database used (e.g., Postgres or MySQL) |

Note that the "pipe separated data" is different than the R pipe (%>%) we will learn about this semester. 

## Moving on to the tutorial

Make sure that the following packages are loaded

```{r}
library(readr) # for reading in data 
```

### Manually creating data

At times it may be necessary to create small datasets manually within R. We will learn two different ways to create small datasets.

#### Vectors

The first type of data we will consider is when you need to make the equivalent of a single variable of values. This is most easily done by create a *vector* in R. Here are a few examples.

```{r}
# different ways to create sequences in R using built in methods.  A lot of these are more quick and dirty compared to if you are trying to import data
x1 <- c(5,1,8,2)
x2 <- 1:12
x3 <- 5:-6
x4 <- c("we","don't", "talk", "about","Bruno")
x5 <- c(x1, x3)
x6 <- c(x1, x4) # converts x1 into a character and combines it with x4
```

If we want to extract individual components out of a vector we do so using commands like the following. Before running each, try to guess what the result will be. Then check your guess by running the code.

```{r}
# different ways to manually extract information from the data that we just generated
x1[1]
x1[4]
x4[1:4]
x4[c(1,3,4,5)]
x1[-1] # negative index tells R to provide everything except for the variable/output that is negated
x1[5]
x1[0] # place holder of length 0, recognizes how things are stored and values appear after

```

#### Tibbles/Data Frames

We can also create a data set directly within `R` with the `tibble()` function in the `tibble` package. This is most useful when we want to make a small reproducible example so that someone else may help with our code. (Note that the `tibble` is automatically loaded when we load the `readr` package/ We will rarely need to load it by itself except in situations where it is the only thing that we are using.)

> A _reproducible example_, or __reprex__, is a chunk of code that we can give to someone else that runs without any outside data. These are used often on StackExchange. The following code chunk is __not__ a reprex because people would not necessarily have the data set `parsedf.csv`.

```{r, message = FALSE, warning = FALSE}
## Hello! How do I get rid of the units from the values in
## my variable `x`? Thanks!
library(here)
test_df <- read_csv(here("data","parsedf.csv"))
head(test_df)
```

We want to post on StackExchange for someone to help us convert a variable from a character vector with units to a numeric vector without units. We want to be able to give any possible helpers a small example data set to work with. For this, we can create our own tiny data set with `tibble()`:

```{r, message = FALSE}
## Hello! How do I get rid of the units from the values in
## my variable `xvar`? Thanks!

# this is a method to create a small data set using the tibble library
library(tibble)
test_df2 <- tibble(xvar = c("20,000 dollars", "40 dollars"),
                   yvar = c(1, 2))
test_df2
```

### Extracting values from a tibble/dataframe

```{r}
# extract a variable by name
test_df2$xvar # stores values as a vector

#extract a variable by location
test_df2[ , 1] # blank in rows spot means take all rows and it will be stored as a tibble

# extract an observation by location
test_df2[2,] # this will pull the entire second row
```


### Importing Files

As we saw in previous lessons, tabular data stored in a CSV files can be read in to R using the `read_csv` function from the `readr` package. If the data is in a clean format already, this is a fairly straightforward process. We will start to investigate how to read in and work with tabular data that is not a simple CSV file.

First, load in the example data sets

```{r}
library(here)

UBSprices <- read_csv(here("data", "UBSprices.csv"))
airlines <- read_csv(here("data", "airline-safety.csv"))
```

Let's take a closer look at the airlines data. Look in your environment panel and expand the airlines object. Notice that it gives you a small amount of information about the variables? Also notice that some of the variables' names have spaces in them. This is generally a bad practice, but is commonplace enough that the R allows a special way to access these variables. Note, you can also check the names of the variables using the `names` function in R.

To access variables with spaces in their names we need to use  a backtick mark, `,  around them. This key is found in the top left corner of the keyboard. Tick marks work a lot like quotes insofar as they open and close and wrap all elements into a single object. The difference is tick marks assume the marked object references a symbol, not a string of characters.

Lets try this out. First use the ticks to extract just the `fatalities 1985_1999` variable.

```{r}
# r will struggle to pull in and recognize variable names with spaces in it
airlines$`fatalities 1985_1999`
```

Note that in some cases, R allows you to be sloppy and you can use quotes instead of ticks. However, it is good practice to try to always use the tick marks when dealing with variable names that contain spaces.

Side note: This can cause some annoyances when sharing code with people that aren't familiar with the tick structure. One way around this is use the `name.repair` option with `make.names` to  "fix" unfriendly names in data.frames. (Using `name_repair = "universal"` is also a useful way to fix these.)

```{r, echo = TRUE, eval = FALSE}
airlines <- read_csv(here("data", "airline-safety.csv"), name_repair = make.names)
```

(Note that the tick mark structure is still useful in other places. We will likely see examples where we will make use of it.)


### Importing tabular files from the web

Suppose you know of a tabular (e.g., csv) file located on the web. Instead of downloading it and then reading it in, you can often just provide the URL as the file. For example, try to import the `slc_towns_active_cases.csv` from [https://github.com/iramler/stlawu_covid](https://github.com/iramler/stlawu_covid){target="_blank"}

```{r}
# copy the url for the raw data from github and place the url in quotes
url = "https://raw.githubusercontent.com/iramler/stlawu_covid/main/slc_towns_active_cases.csv"

active_cases <- read_csv(url)
```

### Exporting Files

Occasionally you may want to save a data set. (e.g., after you have cleaned a messy and/or subsetted a larger dataset) One simple way is to use the `write_csv` function.

Try this now by saving a copy of `test_df2` to your `data` folder. 

```{r}
# exporting data sets by writing new .csv files
# write_csv(x = test_df2, 
          #file = here("data", "test_df2.csv")
               #)
```


Next, lets try a series of exercises to learn about some of the different options available to us in the `readr` (and associated) packages.



### Exercises

1. What are some advantages and disadvantages of using the `Import Dataset` tool in R Studio?

Some of the advantages of using the `Import Dataset` tool is that you don't have to write out the code for it and if you are unsure of the file path, it will define it for you. It also has the ability to change the structure of the data if you need to manipulate it.  Some disadvantages include how it is difficult for others to replicate.  If you were to send this data to other individuals, they probably would not be able to run it because it is specific to the file path you are using. Another disadvantage is that if you use this tool and just put it in the environment, the .Rmd file will not recognize it unless its in the actual .Rmd file. 

2. Read in the "polls" data using read_csv and inspect it. Fix the issue by using the appropriate option in ``read_csv``

```{r}
library(readr)

# read in data set using url, when using this method from github, you need to use the raw data 
polls <- read_csv("https://raw.githubusercontent.com/iramler/stat234/main/notes/data/rcp-polls-messy.csv")
```

Upon inspection of this data, the MoE data field contains numbers but is listed as a character data type. The MoE is referring to the margin of error associated with the poll.  These values are reading in as characters because there is a value that uses two dashes to mark a space with what should be "NA"

To fix this, we can add in a new argument `na = c("","NA","--")`

```{r}
# read in data set using url and redefine the values that are set to NA to also include the -- entry. The col_types() argument is telling the program to look through the data and only pull out the numbers.  Then it will dump everything that is not a number inside of the data.  
polls <- read_csv("https://raw.githubusercontent.com/iramler/stat234/main/notes/data/rcp-polls-messy.csv",
                  na = c("","NA","--"),
                  col_types = list(Sample = col_number()
                                   )
                  
                  )
```

Another issue that we have are spaces in the variable names.  To fix this, we can use the `name_repair = make.names` function to change the structure of the column names.  This standardizes how the names are stored in R

```{r}
# read in data set using url and redefine the values that are set to NA to also include the -- entry. The col_types() argument is telling the program to look through the data and only pull out the numbers.  Then it will dump everything that is not a number inside of the data.  
polls <- read_csv("https://raw.githubusercontent.com/iramler/stat234/main/notes/data/rcp-polls-messy.csv",
                  na = c("","NA","--"),
                  col_types = list(Sample = col_number()),
                  name_repair = make.names
                  
                  )
```



3. Each of the files, `data1.csv`, `data2.csv`, and `data3.xlsx`, have some problems with them when using read_csv. For each file, 
a. import them with ``read_csv``, 

```{r}
# import data1.csv
data1 <- read_csv(here("data", "data1.csv"))

# this does not work because we used the wrong delimiter -> this is a semicolon delimited sheet so we need to specify the deliminator as a semicolon. You can also use read_csv2()

data1_fix1 <- read_delim(here("data", "data1.csv"), 
                    delim = ";"
                    )

# data1_headerchange <- read_csv2(here("data", "data1.csv"), col_names = c("a", "b", "c"))
# this converts the data from numbers to characters so this is not the best way to rename variables

# one way to rename variables that maintains data types
names(data1_fix1) <- c('a', 'b', 'c')
```

```{r}
# import data2.csv
data2 <- read_csv(here("data", "data2.csv"))

# this data set does not have a header so it was imported in a weird way.  If we look at the raw data, it looks like the first row is a row of data as opposed to a header

# Fix 1
data2_fix1 <- read_csv(here("data","data2.csv"), col_names = FALSE)

# Fix 2
data2_fix2 <- read_csv(here("data","data2.csv"), col_names = c("x", "y", "z"))
```

```{r}
# import data 3, to do this we first have to add in the readxl package
library(readxl)


data3 <- read_xlsx(here("data", "data3.xlsx"))

```

