---
title: "Examples with ggplot2"
output: html_document
---

# Brief overview of `ggplots2` package

All `ggplot` functions must have at least three components:

*  **data frame**: data associated with the plot. 
*  **geom**: to determine the type of geometric shape used to display the data, such as line, bar, point, or area. Note that you can use different data in different geoms by 
*  **aes**: to determine how variables in the data are _mapped_ to visual properties (aesthetics) of geoms. This can include x position, y position, color, shape, fill, and size.
  + If the `aes` statement is included in the `ggplot` command, the mapping will be available to any `geom`s used
  + If you need a variable mapped to just a specific `geom`, you can include it in an `aes` mapping for a specific geom. (You can also do something similar with the `data` option if you need to plot multiple data sources on the same display. However, it is generally better to think about structuring your dataset )

# Practice Problems with ggplots2

For this set of exercises we will be using a slightly larger version of the Stat 113 first day survey data. It is in the file `stat113_f18_s19.csv` (and contains two semesters of surveys).

```{r}
library(dplyr)
library(ggplot2)
library(here)
library(readr)

stat113 <- read_csv(here("data", "stat113_f18_s19.csv"))
```

## Univariate Analyses

We will begin with a series of basic displays for univariate analyses.

* Investigate the distribution of the amount of TV watched (in hours per week) by Stat 113 students.
```{r}
base_plot <- ggplot(stat113, mapping = aes(x = TV))

```

* Use another display to investigate the distribution of the amount of TV watched (in hours per week) by Stat 113 students.
```{r}
base_plot+
     geom_density(color = "blue",
                  fill = "red",
                  alpha = 0.5)
```

* Combine the geoms from the previous part onto the same plot.
```{r}
# not working because the density and histogram scales are different
base_plot + geom_histogram() + geom_density(
     color = "blue")
```

```{r}
# force the histogram to be on the density scale
base_plot + 
     geom_histogram(aes(y = after_stat(density)))+
     geom_density()
```

```{r}
# force the density to be on the count scale
base_plot +
     geom_histogram(bins = 50)+
     geom_density(aes(y = after_stat(count)))
```



* Pick your favorite chart of the three and play around with a few options such as color, fill, linetype, and alpha. 



* Pick another numeric variable to explore. Play around with themes, coordinate systems, and labels.
```{r}
ggplot(data = stat113, aes(Exercise))+
     geom_density(fill = "dodger blue",
                  alpha = 0.25,
                  )+
     theme_bw()+ 
     coord_flip()+
     labs(caption = "Figure 1: Distribution of ...",
          x = "Exercise per week (hours)")

#coord_polar is a way to plot circular data (ex: wind direction, time, etc.)
     
```


* Pick a categorical variable and make a bar chart.
```{r}
award_plot <-
stat113 %>%
     filter(!is.na(Award)) %>%
     ggplot(aes(x = Award, fill = Award))+
     geom_bar()
```

```{r}
# saving plots
ggsave(filename = "stat113_award_plot.jpg",
       plot = award_plot,
       width = 7,
       height = 4)
```

## Multivariate Analyses

We will investigate a few research questions involving two variables and produce graphics of the following. (Note: You may need to use different sets of variables for some of them.)

* Side-by-side boxplots and/or violin plots
```{r}
# make side-by-side boxplots
stat113%>% 
  filter(!is.na(Sport)) %>%
  ggplot(aes(x = Sport, 
             y = GPA, 
             fill = Sport))+
  geom_violin()+
  scale_fill_manual(
    values = c("skyblue", "royalblue"),
    name = "Student\nAthlete" # \n is used to insert a line break in text
  ) +
  labs(title = "title of plot",
       caption = "caption of plot")+
  coord_cartesian(ylim = c(0,4.1), expand = FALSE) # NA can also be used to define the upper bound based on how big the data gets
```

* Scatterplot
```{r}
# create a scatter plot 
stat113 %>%
  ggplot(aes(x = Exercise, y= Pulse ))+
  geom_point()
```

* Scatterplot + Smoother
```{r}
# create a scatter plot 
stat113 %>%
  ggplot(aes(x = Exercise, y= Pulse ))+
  geom_point()+
  geom_smooth()
```
* Scatterplot + Linear Smoother
```{r}
# create a scatter plot 
stat113 %>%
  ggplot(aes(x = Exercise, y= Pulse ))+
  geom_point()+
  geom_smooth(method = lm)
```
```{r}
# create a scatter plot 
stat113 %>%
  filter(!is.na(Tattoo)) %>%
  ggplot(aes(x = TV, y= GPA, 
             color = Tattoo, fill = Tattoo, shape = Tattoo))+
  geom_point()+
  geom_smooth()+
  geom_smooth(method = lm) +
  facet_wrap(vars(Tattoo), scales = "free") +
  theme_bw()

# standard errors are controlled by the fill command
```

* Stacked Bar Chart
```{r}
stat113 %>%
  filter(!is.na(Award), !is.na(Sport)) %>%
  ggplot(aes(x = Award, fill = Sport))+
  geom_bar(position = "stack")
```

```{r}
# cluster bar chart
stat113 %>%
  filter(!is.na(Award), !is.na(Sport)) %>%
  ggplot(aes(x = Award, fill = Sport))+
  geom_bar(position = "dodge")
```
```{r}
# filled bar chart, changes the y-axis to be a proportion
stat113 %>%
  filter(!is.na(Award), !is.na(Sport)) %>%
  ggplot(aes(x = Sport, fill = Award))+
  geom_bar(position = "fill")+
  labs(y= "Proportion") +
  scale_fill_viridis_d() +
  #scale_fill_brewer(palette = "Accent")+
  theme_bw()
```
* Faceted Density plots
```{r}
stat113 %>%
  ggplot(aes(x = Class, y = GPA, group = Class)) +
  
  geom_violin(aes(fill = Class))+
  
 
  geom_boxplot()+
  
  facet_wrap(vars(Term))+
  
  theme_bw() +
  coord_cartesian(ylim = c(0,4.1), expand = FALSE)+
  labs(x = "Stat 113 Section")

```

After getting a "basic plot" constructed for each, investigate options to customize and "clean up" the plots. Try to make them look nice.


* Recreate "example.png" (from the T drive)

* Use Class as a factor (categorical variable) instead of a numerical variable

## Further Practice

The document `06-data_visualization` contains more detailed notes related to the grammar of graphics (using `ggplots2`). Please note that a set of Exercises will have you working through those from that document. You might as well get started on them early.