---
title: "Project 5 - Cleaning data"
output: html_notebook
---

# Overview:
In this project, we are playing around with data on the demographics for different movies.  For this project, I chose movies with Ben Stiller in them.  The three movies are "Dodgeball", "The Secret Life of Walter Mitty", and "A Night at the Museum".

```{r}
# clear the environment and import the data
rm(list = ls())
library(tidyverse)
library(here)

waltermitty <- read_csv(here("data", "waltermitty_imbddemographics.csv"))
museum <- read_csv(here("data","nightatmuseum_imbddemographics.csv"))
dodgeball <- read_csv(here("data","dodgeball_imbddemographics.csv"))

```

# Cleaning up the data
Now that the data has been imported, we need to clean it up. To do this, I will start with the demographics data from "The Secret Life of Walter Mitty"

```{r}
waltermitty_tidy <-
     waltermitty %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

```{r}
museum_tidy <-
     museum %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

```{r}
dodgeball_tidy <-
     dodgeball %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

# Analysing the data
Now that the data is in a clean format, I will analyse the data to extract any trends and/or relationships between the movies.

```{r}
# graphing the ratings by demographic
waltermitty_tidy %>%
     ggplot(aes(x=Rating, fill = Demographic)) +
```

