---
title: "Project 5 - Cleaning data"
output: html_notebook
---

# Overview:
In this project, we are playing around with data on the demographics for different movies.  For this project, I chose movies with Ben Stiller in them.  The three movies are "Dodgeball", "The Secret Life of Walter Mitty", and "A Night at the Museum".

```{r}
# clear the environment and import the data
rm(list = ls())
library(tidyverse)
library(here)

waltermitty <- read_csv(here("data", "waltermitty_imbddemographics.csv"))
museum <- read_csv(here("data","nightatmuseum_imbddemographics.csv"))
dodgeball <- read_csv(here("data","dodgeball_imbddemographics.csv"))

```

# Cleaning up the data
Now that the data has been imported, we need to clean it up. To do this, I will start with the demographics data from "The Secret Life of Walter Mitty"

```{r}
waltermitty_tidy <-
     waltermitty %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

```{r}
museum_tidy <-
     museum %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

```{r}
dodgeball_tidy <-
     dodgeball %>%
     rename(., "Demographic" = 1) %>%
     pivot_longer(cols = c(2:6),
                  names_to = "Age_range",
                  values_to = "Rating") %>%
     mutate(Rating = str_squish(Rating)) %>%
     separate(col = "Rating",
              sep = " ",
              into = c("Rating", "Number_of_Votes"))
     
```

Join all of the movies together:
```{r}
movies <- bind_rows(museum_tidy,
                   dodgeball_tidy,
                   waltermitty_tidy,
                   .id = "Movie")

# convert the Rating and number of votes values to numbers
movies <- movies %>%
        mutate( 
                Rating = parse_number(Rating),
                Number_of_Votes = parse_number(Number_of_Votes),
                Demographic = as.factor(Demographic),
                Movie = as.factor(Movie),
                Age_range = as.factor(Age_range),
                Movie = fct_recode(Movie, "The Secret Life of Walter Mitty"= "3", "Dodgeball" = "2", "Night at the Museum" = "1")
                )
       
            
```


# Analysing the data
Now that the data is in a clean format, I will analyse the data to extract any trends and/or relationships between the movies.

```{r}
# graphing the ratings by demographic
movies %>%
     ggplot(aes(x= Movie, y =Rating, fill = Demographic)) +
        geom_boxplot()+
        theme_bw() +
        labs(y = "IMBd Rating")
        
        
```

```{r}
movies %>%
        ggplot(aes(x = Demographic, y = Rating, fill = Movie)) +
        geom_boxplot()+
        labs(y = "IMBd Rating")+
        theme_bw()
```
```{r}
movies %>%
        filter(Age_range != "All Ages") %>%
        ggplot(aes(x = Age_range, y = Number_of_Votes)) +
        geom_boxplot()+
        geom_point(aes(group = Demographic, color = Demographic), position = "jitter")+
        labs(x = "Age Range", y = "Number of Votes") +
        theme_bw()
```
Outline for presentation:
- Introduce the movies and their relationships to one another (all are movies with Ben Stiller)
- Talk about IMBd and how the ratings work
- Show number of votes by age range boxplt



